---
title: "Getting started with cleanplots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cleanplots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5, dpi = 96,
  message = FALSE, warning = FALSE
)
has_mfx <- requireNamespace("marginaleffects", quietly = TRUE)
has_msm <- requireNamespace("modelsummary", quietly = TRUE)
has_sim <- tryCatch({
  sim <- readRDS(url(
    "https://raw.githubusercontent.com/tdmize/data/master/data/sim_lifesat.rds",
    open = "rb"))
  TRUE
}, error = function(e) FALSE)
```

cleanplots provides publication-ready defaults for ggplot2. The goal is
professional-looking figures with strong data visualization and
accessibility defaults, with no per-plot effort: a color palette that
is **colorblind-friendly** and remains **distinguishable when printed
in black & white**, matching marker shapes and line patterns that keep
groups distinct across multiple visual channels, a clean theme, and
consistent figure sizing. A companion
[cleanplots scheme for Stata](https://www.trentonmize.com/software/cleanplots)
uses the same design, so figures made in R and Stata look like
siblings.

The package is **layered**: use as little or as much as you want. This
vignette walks from the lightest touch (colors only) to the full setup
(one call that changes everything).

```{r setup}
library(ggplot2)
library(cleanplots)
```

## 1. Colors only

If all you want is the cleanplots colors -- leaving the theme, sizes,
and everything else exactly as they were -- add
`scale_color_cleanplots()` (or `scale_fill_cleanplots()`) to a plot,
just like any other palette package:

```{r colors-only}
ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point() +
  scale_color_cleanplots() +
  theme_minimal()
```

One recommendation: if you are not using the full cleanplots setup
(`theme_cleanplots()` or `cleanplots_defaults()`, below), add
`theme_minimal()` to your plots -- the colors and markers are much
easier to see against a white background than against ggplot2's
default gray. The examples in this vignette do this.

Both scales take the same options:

* `palette`: `"default"` (the main colors) or `"bars"` (softer versions
  for bars, areas, and pies -- more on this below)
* `order`: use the colors in a different order, e.g.
  `order = c(7, 1, 2)` starts with navy
* `reverse`: reverse the palette

```{r colors-order}
ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point() +
  scale_color_cleanplots(order = c(7, 1, 2)) +
  theme_minimal()
```

To use individual colors manually, `cleanplots_colors()` returns hex
codes by name:

```{r colors-extract}
cleanplots_colors()
cleanplots_colors("red", "navy")
```

## 2. The two palettes

The **main palette** is used for markers, lines, and confidence
intervals. It alternates darker and lighter colors so that the first
several groups remain distinguishable even when printed in black &
white, and it contains no red--green pair (red-green colorblindness is
the most common form, affecting roughly 8% of men).

The **bar palette** contains softer versions of the same colors. Bars,
areas, and pie slices use far more ink than points and lines, so
full-strength colors overwhelm; cleanplots uses gentler fills for
these elements:

```{r bars-palette}
titanic <- aggregate(Freq ~ Class + Sex, data = as.data.frame(Titanic), sum)

ggplot(titanic, aes(Sex, Freq, fill = Class)) +
  geom_col(position = "dodge") +
  scale_fill_cleanplots(palette = "bars") +
  labs(x = NULL, y = "Passengers and crew", fill = "Class") +
  theme_minimal()
```

## 3. Shapes and line patterns

Color alone reliably distinguishes about 4--5 groups. Beyond that --
and for black & white printing and colorblind readers -- cleanplots
varies **multiple visual channels at once**, so that no group pair
ever depends on a single cue:

* **Color and lightness**: the palette alternates dark and light.
* **Marker shape and fill**: `scale_shape_cleanplots()` gives hollow
  shapes (circle, square, triangle, diamond) to the dark colors and
  solid shapes to the light colors.
* **Line pattern**: `scale_linetype_cleanplots()` assigns solid,
  solid, dashed, dashed, shortdash, shortdash, longdash, longdash.

The assignments are coordinated: any two groups that share a shape or
a line pattern always differ strongly in lightness, so every pair of
groups is separated by at least two independent channels.

```{r shapes}
ggplot(mpg, aes(displ, hwy, color = class, shape = class)) +
  geom_point(size = 2, stroke = 0.7) +
  scale_color_cleanplots() +
  scale_shape_cleanplots() +
  theme_minimal()
```

```{r linetypes}
ggplot(economics_long, aes(date, value01, color = variable, linetype = variable)) +
  geom_line(linewidth = 0.75) +
  scale_color_cleanplots() +
  scale_linetype_cleanplots() +
  theme_minimal()
```

## 4. The theme

`theme_cleanplots()` applies the cleanplots layout: white background,
no plot border, light gray axis lines, dotted gridlines, a frameless
legend at the right, and black-outlined facet strips. Like any ggplot2
theme, add it per plot:

```{r theme}
ggplot(mpg, aes(displ, hwy, color = class, shape = class)) +
  geom_point(size = 2, stroke = 0.7) +
  scale_color_cleanplots() +
  scale_shape_cleanplots() +
  theme_cleanplots()
```

## 5. The full setup: `cleanplots_defaults()`

Adding scales and themes to every plot gets repetitive. One call at
the top of your script applies the complete cleanplots setup for the
session:

```{r defaults}
cleanplots_defaults()
```

After this, with **no scales or theme added**:

* every plot uses `theme_cleanplots()`;
* discrete `color` aesthetics use the main palette, and discrete
  `fill` aesthetics use the softer bar palette (bars and areas
  automatically get the gentler colors);
* points are larger with heavier outlines (`size = 2`,
  `stroke = 0.7`), so the hollow marker shapes are clearly visible;
* lines are thicker (`linewidth = 0.75`) for `geom_line()`,
  `geom_path()`, `geom_step()`, `geom_density()`, `geom_function()`,
  and `geom_smooth()`; error bars, linereanges, and pointranges use
  80% of that;
* `geom_smooth()` fit lines are cleanplots red with a light gray
  confidence band, instead of ggplot2's blue.

```{r defaults-demo}
ggplot(mpg, aes(displ, hwy, color = class, shape = class)) +
  geom_point() +
  scale_shape_cleanplots()
```

```{r defaults-smooth}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(color = cleanplots_colors("gray")) +
  geom_smooth(method = "loess", formula = y ~ x)
```

All settings are arguments if you prefer different sizes:

```{r defaults-args, eval = FALSE}
cleanplots_defaults(base_size = 12, point_size = 2,
                    point_stroke = 0.7, line_width = 0.75,
                    smooth_color = "#D50000")
```

The default text size (12) is chosen to match the body text of an
academic article when the figure is saved at the recommended size (see
the next section): if you can read the article, you can read the
graph. For contexts that need bigger text -- presentations and
lectures especially -- increase it:

```{r defaults-slides, eval = FALSE}
cleanplots_defaults(base_size = 16)   # presentations and lectures
```

Everything remains overridable per plot: an explicit scale, theme, or
aesthetic always wins. For example, to use the main palette for a fill
instead of the automatic bar palette:

```{r override}
ggplot(titanic, aes(Sex, Freq, fill = Class)) +
  geom_col(position = "dodge") +
  scale_fill_cleanplots(palette = "default") +
  labs(x = NULL, y = "Passengers and crew", fill = "Class")
```

**To restore ggplot2's defaults**: `cleanplots_defaults()` sets
session-wide state, which persists until you restart R. To reset
manually:

```{r undo, eval = FALSE}
theme_set(theme_gray())
options(ggplot2.discrete.colour = NULL, ggplot2.discrete.fill = NULL)
update_geom_defaults("point", list(size = 1.5, stroke = 0.5))
```

## 6. Saving figures

An important quirk of ggplot2: a plot has no intrinsic size. Text,
markers, and lines are physical units (points and millimeters), and
`ggsave()` **without explicit dimensions saves at whatever size your
plot window happens to be** -- so the same code produces figures with
different proportions on different days, machines, and window layouts.

The fix is to always save at explicit dimensions. `cleanplots_save()`
does this with tuned defaults -- 7 x 5 inches at 300 dpi:

```{r save, eval = FALSE}
p <- ggplot(mpg, aes(displ, hwy, color = drv)) + geom_point()
cleanplots_save("my-figure.png", p)
```

Why 7 x 5? It matches the default figure size of R Markdown (HTML)
documents, and a 7-inch figure placed at the 6.5-inch text width of a
US-letter manuscript scales the 12-point default text to about 11
points -- comparable to the article's body text. The file extension
sets the format, so `.pdf`, `.tiff`, or `.eps` for journal submission
systems work directly. Override any default as needed:

```{r save2, eval = FALSE}
cleanplots_save("slide-figure.png", p, width = 10, height = 5.6)
```

The same principle applies to other ways of saving:

* **`ggsave()` directly**: always pass `width`, `height`, and `dpi`.
* **RStudio's Export button**: type explicit dimensions in the dialog
  (e.g., 2100 x 1500 pixels = 7 x 5 inches at 300 dpi) rather than
  accepting the window size.
* **R Markdown / Quarto**: figures are sized by chunk options, which
  is already fixed-size by construction. To match `cleanplots_save()`:
  `knitr::opts_chunk$set(fig.width = 7, fig.height = 5)` (or
  `fig-width`/`fig-height` in Quarto). Note that `pdf_document`
  defaults to 6.5 x 4.5 -- article text width -- which also works
  well.
* **Base devices** (`png()` ... `print(p)` ... `dev.off()`): pass
  dimensions to the device call.

Avoid copy-pasting figures from the plot window into Word or
PowerPoint: the result is window-sized and screen-resolution.

## 7. Plots of predictions and marginal effects

cleanplots works out of the box with the
[marginaleffects](https://marginaleffects.com) and
[modelsummary](https://modelsummary.com) packages. With
`cleanplots_defaults()` set, their plots pick up the colors and theme
automatically.

The examples below use data on life satisfaction across ages:

```{r sim-load, eval = FALSE}
sim <- readRDS(url(
  "https://raw.githubusercontent.com/tdmize/data/master/data/sim_lifesat.rds",
  open = "rb"))
```

A coefficient plot of nested models with
`modelsummary::modelplot()`:

```{r coefplot, eval = has_msm && has_sim}
mod1 <- lm(lifesat ~ age + educ, data = sim)
mod2 <- lm(lifesat ~ age + educ + married, data = sim)

modelsummary::modelplot(list("Baseline" = mod1, "+ Marriage" = mod2),
                        coef_omit = "Intercept") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  labs(x = "Coefficient estimates and 95% confidence intervals")
```

Adjusted predictions for a nominal predictor with
`marginaleffects::plot_predictions()`:

```{r preds-nominal, eval = has_mfx && has_sim}
mod <- lm(lifesat ~ poly(age, 2) * married + educ, data = sim)

marginaleffects::plot_predictions(mod, condition = c("educ", "married")) +
  labs(y = "Predicted life satisfaction")
```

Predictions across a continuous predictor by group. One note here:
`plot_predictions()` draws its confidence-interval ribbons with heavy
transparency (alpha = 0.1), which is tuned for saturated colors. Add
`scale_fill_cleanplots(palette = "default")` so the ribbons use the
full-strength palette rather than the soft bar colors:

```{r preds-continuous, eval = has_mfx && has_sim}
marginaleffects::plot_predictions(mod, condition = c("age", "married")) +
  scale_fill_cleanplots(palette = "default") +
  labs(y = "Predicted life satisfaction")
```

For full control of the ribbons (or anything else), use
`plot_predictions(..., draw = FALSE)` to get the plotting data and
build the figure yourself:

```{r preds-draw-false, eval = has_mfx && has_sim}
pr <- marginaleffects::plot_predictions(
  mod, condition = c("age", "married"), draw = FALSE)

ggplot(pr, aes(age, estimate, color = married, fill = married)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = .35, color = NA) +
  geom_line() +
  labs(y = "Predicted life satisfaction")
```

And comparisons -- here, the marriage gap in life satisfaction across
ages -- with `marginaleffects::plot_comparisons()`:

```{r comparisons, eval = has_mfx && has_sim}
marginaleffects::plot_comparisons(mod, variables = "married",
                                  condition = "age") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(y = "Difference in predicted life satisfaction\n(Married - Not married)")
```

## 8. For Stata users

A companion cleanplots scheme for Stata shares this package's colors,
marker symbols, line patterns, and layout -- the hex values in both
languages are computed from the same definitions, so colors match
exactly. The mapping:

| Stata | R |
|---|---|
| `set scheme cleanplots` | `cleanplots_defaults()` |
| `p1`--`p10` colors | `scale_color_cleanplots()` |
| `p1bar`--`p10bar` colors at `intensity bar 70` | `scale_fill_cleanplots(palette = "bars")` |
| marker symbols (`symbol p1` ...) | `scale_shape_cleanplots()` |
| line patterns (`linepattern p1line` ...) | `scale_linetype_cleanplots()` |
| scheme layout settings | `theme_cleanplots()` |
| graph export sizing | `cleanplots_save()` |

The Stata scheme is available at
<https://www.trentonmize.com/software/cleanplots>.
